Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2025-01-10 manuscript/manuscript.tex
Minor edits for readability. Add "field usage by Agency" table

\begin{abstract}
In the early 21st century, the open data movement began to transform 
societies and governments by promoting transparency,
innovation, and public engagement. The City of New York (NYC) has been at
the forefront of this movement since the enactment of the Open 
Data Law in 2012, leading to the creation of the NYC Open Data
portal. This portal currently hosts 2,700 datasets,
serving as a crucial resource for research across various domains, 
including health, urban development, and transportation. While the
open data movement has made vase amounts of data accessible, the
challenge of ensuring data quality and usability through effective
curation remain insufficiently addressed in the literature..
This paper examines these challenges through a
case study of the NYC 311 Service Request dataset, focusing on issues 
of data validity, consistency, and curation efficiency. Based on 
insights from this case study, we propose a set of data curation 
principles tailored for government-released open data aiming
to enhance data management practices and ensure 
the ongoing utility of open data.
\end{abstract}

@@

Today, the NYC 311 data system manages over 3 million service
requests (SRs) per year. This data is publicly accessible 
through the NYC Open Data Portal, which provides 
tools for querying, grouping, aggregating,
geo-mapping, visualizing, and exporting results. The
NYC Office of Technology and Innovation (OTI) provides technical support for 
the 311 application, the open data infrastructure, mobile \& web applications, 
and the public-facing data portal. Input data for the 311 system 
is sourced from 16 different NYC Agencies. Some of these Agencies
utilize the core 311 software, while others transfer data to the
311 Open Data Warehouse from their in-house systems.

@@

\begin{itemize}[left=1.5em]
\item The dataset is 3.7 GB in size and contains 
  over 6.4 million rows. It was exported as a CSV file.

@@

While one can likely infer the meaning of these ``computed'' 
fields, uncertainty remains as to their derivation and 
usage. These six fields are not included in
the 311 SR Data Dictionary. Additionally, our analysis 
revealed that these six ``computed'' fields have 
significant data validity issues, leading us to categorize these six fields 
as experimental and \textit{not for public use}. An update 
to the Data Dictionary would seem appropriate. Accordingly, 
we will not address these fields in our analysis.

\paragraph{Representation of Missing data:} We found that the 
311 SR dataset exhibited inconsistent approaches when 
presenting missing data. We discovered many different 
representations of missing data: nulls, spaces, ``NA'', ``N/A'', 
and ``<NA>''. This variety of representations complicates programming
efforts, and increases the likelihood of introducing errors 
in data analysis (typically by search for or excluding missing fields 
and not aware of the various representations of such). A more consistent 
approach to treating missing data should be employed to address this issue.

@@

\subsection{Missing Data by Field}
\label{sec:blanks}
Understanding the absence of data by field greatly aids when 
conducting analysis. For example, determining if SRs were closed 
before or after their \texttt{due\_date}, you would be challenged as 99.6\% of the
\texttt{due\_date} fields are blank, and are only used by one Agency, DSNY. Blank 
(or N/A values) they appear to divide into three groups:

@@

The Mostly Empty category includes:
\texttt{taxi\_company\_borough}, \texttt{due\_date}
\texttt{road\_ramp}, and \texttt{bridge\_highway\_name}. The 
Partially Empty category includes:
\texttt{location\_type}, \texttt{landmark}, 
and \texttt{cross\_street\_1 \& \_2}. And the Few/None Empty 
category includes \texttt{created\_date}, \texttt{complaint\_type},
\texttt{agency}, and \texttt{status}. Of particular note is that 
several of the Mostly Empty fields are used by only one or 
two Agencies (See: \autoref{tab:field-usage-summary}).

\begin{table}[htbp]
    \centering
    \caption{Fields used by only one or two Agencies with limited non-blank entires}
    \label{tab:field-usage-summary}
    \begin{tabular}{l r r p{4cm}}
        \toprule
        \textbf{Field} & \textbf{\# of entries} & \textbf{\% non-blank} & \textbf{Agencies using} \\
        \midrule
        \texttt{taxi\_company\_borough} & 3,567  & 0.06\% & TLC \\
        \texttt{vehicle\_type}           & 18,642 & 0.29\% & TLC, NYPD \\
        \texttt{due\_date}               & 24,414 & 0.38\% & DSNY \\
        \texttt{taxi\_pick\_up\_location} & 64,954 & 1.02\% & TLC \\
        \bottomrule
    \end{tabular}
\end{table}

For some analysis efforts, it may be prudent to inquire 
 as to why some fields are mostly/almost always blank. 
Figure~\ref{fig:blank_fields} is a graphic depiction of the \%
empty (blank \& N/As) for each field. It illustrates the grouping into 
Mostly Empty, Partially Empty, and Few/None Empty segments.

@@

@@

However, some fields lacked compliance
to a domain of allowable values, including:

@@

\begin{itemize}[left=1.5em]
    \item 12,450 SRs with a \texttt{closed\_date} that occurs before the 
    \texttt{created\_date} resulting a nonsensical ``negative duration''. 
    \item Eight SRs with \texttt{created\_date(s)} or \texttt{closed\_date(s)} in 
    the distant past, specifically ``1900-01-01''. Despite the small number of 
    these entries, such extreme values can greatly distort analytical conclusions.
    \item 163,720 SRs with identical \texttt{created\_date(s)} and \texttt{closed\_date(s)} 
    to the second, creating a nonsensical ``zero duration''. 
    \item 247,000 SRs closed and/or created exactly at midnight, to the second.
\end{itemize}

@@

\begin{figure}[tbp]
    \centering
    % First subfigure
    \begin{subfigure}[t]{0.495\textwidth} % Adjust width to fit side by side
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_created_on_the_hour.pdf}
        \caption{SRs created exactly on the hour}
        \label{fig:busiestcreated}
    \end{subfigure}
    \hfill % Add horizontal space between subfigures
    % Second subfigure
    \begin{subfigure}[t]{0.495\textwidth} % Adjust width to fit side by side
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_closed_on_the_hour.pdf}
        \caption{SRs closed exactly on the hour}
        \label{fig:busiestclosed}
    \end{subfigure}
    % Main caption
    \caption{Comparison of 311 SRs created \& closed by hour-of-the-day during 2022--2023.}
    \label{fig:exacthours}
\end{figure}

@@

\begin{quote}
\textit{Note: The authors attempted extensive standardization of street names  
in an attempt to match what were obvious identical streets, deemed unequal 
due only to spelling and naming conventions. Those efforts were intended to match, for example, 
"KANE PL" and "KANE PLACE" as well as "FLATLANDS 7 ST" and "FLATLANDS 7 STREET". However, these 
efforts were  abandoned after they increased street name matches by less than 1\%.)}
\end{quote}

@@

    \item \texttt{cross\_street\_1/2} and \texttt{intersection\_street\_1/2}: 
    These two street pairs exhibit an 88\% match. Although we cannot 
    confirm the accuracy of either of these street pairs, we recommend deleting the 
    \texttt{intersection\_street\_1/2} fields while acknowledging 
    the concomitant data loss.
\end{itemize}

Additionally, certain fields may not be useful for broad analysis
due to their sparse data population. This especially applies to the fields below, 
which are predominantly used by the TLC and are 99\% (or greater) blank.

\begin{quote}
\textit{[We should note that even a field that is 99\% 
blank will still contain 64,000 non-blank rows in this 
large (6.4 million row) dataset.}; non-trivial.]
\end{quote}

@@

The 311 SR Data Dictionary would greatly benefit from an update. We
found several discrepancies between the Data Dictionary and the actual
data. Such errors could potentially mislead analytic efforts. Examples include:

@@

\section*{Supplementary Material}
The NYC 311 SR data and code used to produce the results reported in
this paper are available at \url{https://github.com/jun-yan/nyc311clean}.

>>>>

\begin{abstract}
In the early 21st century, the open data movement began to transform 
societies and governments by promoting transparency,
innovation, and public engagement. The City of New York (NYC) has been at
the forefront of this movement since the enactment of the Open 
Data Law in 2012, leading to the creation of the NYC Open Data
portal. This portal currently hosts 2,700 datasets,
serving as a crucial resource for research across various domains, 
including health, urban development, and transportation. While the
open data movement has made vase amounts of data accessible, the
challenge of ensuring data quality and usability through effective
curation remain insufficiently addressed in the literature..
This paper examines these challenges through a
case study of the NYC 311 Service Request dataset, focusing on  
data validity, consistency, and curation efficiency. Based on 
insights from this case study, we propose a set of data curation 
principles tailored for government-released open data aiming
to enhance data management practices and ensure 
the ongoing utility of open data.
\end{abstract}

@@

Today, the NYC 311 data system manages over 3 million service
requests (SRs) per year. This data is publicly accessible 
through the NYC Open Data Portal, which provides 
tools for querying, grouping, aggregating,
visualizing, and exporting results. The NYC Office of Technology and Innovation (OTI) 
provides technical support for the 311 application, 
the open data infrastructure, mobile \& web applications, 
and the public-facing data portal. Input data for the 311 system 
is sourced from 16 different NYC Agencies. Some of these Agencies
utilize the core 311 software, while others transfer data to the
311 Open Data Warehouse from their in-house systems.

@@

\begin{itemize}[left=1.5em]
\item The dataset is 3.7 GB in size and contains 
  over 6.4 million rows. It exports as a CSV file.

@@

While one can likely infer the meaning of these \textit{computed} 
fields, uncertainty remains as to their derivation and 
usage. These six fields are not included in
the 311 SR Data Dictionary. Additionally, our analysis 
revealed that these six ``computed'' fields have 
significant data validity issues, leading us to categorize these six fields 
as experimental and \textit{not for public use}. An update 
to the Data Dictionary would seem appropriate. Accordingly, 
we omit these fields in our analysis.

\paragraph{Representation of Missing data:} We found that the 
311 SR dataset exhibited inconsistent approaches when 
presenting missing data. We discovered different 
representations of missing data: nulls, spaces, ``NA'', ``N/A'', 
and ``<NA>''. This variety of representations complicates programming
efforts and increases the likelihood of introducing errors 
in data analysis (typically when selecting data for blank, NA, or missing fields). A more consistent 
approach to treating missing data should be employed to address this issue.

@@

\subsection{Missing Data by Field}
\label{sec:blanks}
Understanding the absence of data by field greatly aids analysis. 
For example, determining if SRs were closed 
before their \texttt{due\_date}, would be challenging as 99.6\% of the
\texttt{due\_date} entries are blank as only DSNY appears to use this field.
Blank (or N/A values) divide into three groups:

@@

\vspace{\baselineskip}
\noindent
The Mostly Empty category includes:
\texttt{taxi\_company\_borough}, 
\texttt{bridge\_highway\_name},
\texttt{due\_date}, and \texttt{road\_ramp}. 
The Partially Empty category includes:
\texttt{cross\_street\_1 \& \_2},
\texttt{location\_type}, and texttt{landmark}. And 
the Few/None Empty category includes \texttt{created\_date}, 
\texttt{complaint\_type},
\texttt{agency}, 
and \texttt{status}. 
Of particular note is that several of the Mostly Empty fields are used by only one or 
two Agencies (See: \autoref{tab:field-usage-summary}).

\begin{table}[tbp]
    \centering
    \caption{Fields used by only one or two Agencies with limited non-blank entires}
    \label{tab:field-usage-summary}
    \small
    \begin{tabular}{l r r p{4cm}}
        \toprule
        \textbf{Field} & \textbf{\# of entries} & \textbf{\% non-blank} & \textbf{Agencies using} \\
        \midrule
        \texttt{taxi\_company\_borough} & 3,567  & 0.06\% & TLC \\
        \texttt{vehicle\_type}           & 18,642 & 0.29\% & TLC, NYPD \\
        \texttt{due\_date}               & 24,414 & 0.38\% & DSNY \\
        \texttt{taxi\_pick\_up\_location} & 64,954 & 1.02\% & TLC \\
        \bottomrule
    \end{tabular}
\end{table}

% Force a blank line and remove indentation
\vspace{\baselineskip}
\noindent
For some analysis efforts, it may be prudent to inquire 
 as to why some fields are mostly/almost always blank. 
Figure~\ref{fig:blank_fields} is a graphic depiction of the \%
empty (blank \& N/As) for each field. It illustrates the grouping into 
Mostly Empty, Partially Empty, and Few/None Empty segments.

@@

\vspace{\baselineskip}
\noindent
The accomanying R program also produces a CSV file showing Field usage by Agency. 
Table~\ref{tab:field-usage-by-agency} contains a sample of Field usage by select Agencies.

\begin{table}[tbp]
\centering
\caption{Sample Field Usage by select Agencies}
\label{tab:field-usage-by-agency}
\small
\begin{tabular}{lrrrrrrr}
\toprule
\textbf{\textbf{Field}} & \textbf{DEP} & \textbf{DHS} & \textbf{DOB} & \textbf{DOT} & \textbf{DPR} & \textbf{DSNY} & \textbf{TLC} \\
\midrule
\texttt{landmark}                  & 2935     & 70185    & 282      & 135150   & 254218   & 424284   & 63133  \\
\texttt{facility\_type}            & 175549   & 0        & 51575    & 159667   & 0        & 66012    & 0      \\
\texttt{due\_date}                 & 0        & 0        & 0        & 0        & 0        & 32097    & 0      \\
\texttt{community\_board}          & 361715   & 99613    & 203715   & 401143   & 266201   & 608188   & 73601  \\
\texttt{vehicle\_type}             & 0        & 0        & 0        & 0        & 0        & 0        & 1696   \\
\texttt{taxi\_company\_borough}    & 0        & 0        & 0        & 0        & 0        & 0        & 4025   \\
\texttt{taxi\_pick\_up\_location}  & 0        & 0        & 0        & 0        & 0        & 0        & 72558  \\
\texttt{bridge\_highway\_name}     & 15       & 21012    & 0        & 7451     & 0        & 613      & 977    \\
\texttt{bridge\_highway\_direction} & 15      & 8973     & 0        & 7450     & 0        & 614      & 1006   \\
\texttt{road\_ramp}                & 15       & 2588     & 0        & 7450     & 0        & 613      & 999    \\
\texttt{bridge\_highway\_segment}  & 15       & 21013    & 0        & 7450     & 0        & 613      & 987    \\
\bottomrule
\end{tabular}
\end{table}

@@

\vspace{\baselineskip}
\noindent
However, some fields lacked compliance
to a domain of allowable values, including:

@@

\begin{itemize}[left=1.5em]
    \item 12,450 SRs with a \texttt{closed\_date} before \texttt{created\_date} creating a nonsensical ``negative duration''. 
    \item Eight SRs with \texttt{created\_date(s)} or \texttt{closed\_date(s)} in 
    the distant past, specifically  \textit{1900-01-01}. Despite few of 
    these entries, such extreme values can greatly distort analytical efforts. 
    \item 163,720 SRs with identical \texttt{created\_date(s)} and \texttt{closed\_date(s)} 
    to the second, creating a nonsensical ``zero duration''. 
    \item 247,000 SRs closed and/or created exactly at midnight, to the second.
\end{itemize}

@@

\begin{figure}[tbp]
    \centering
    % First subfigure
    \begin{subfigure}[t]{0.495\textwidth} % Adjust width to fit side by side
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_created_on_the_hour.pdf}
        \caption{SRs created exactly on the hour}
        \label{fig:busiestcreated}
    \end{subfigure}
    \hfill % Add horizontal space between subfigures
    % Second subfigure
    \begin{subfigure}[t]{0.495\textwidth} % Adjust width to fit side by side
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_closed_on_the_hour.pdf}
        \caption{SRs closed exactly on the hour}
        \label{fig:busiestclosed}
    \end{subfigure}
    % Main caption
    \caption{311 SRs Created \& Closed exactly on the hour (hh:00:00) for 2022--2023.}
    \label{fig:exacthours}
\end{figure}

@@

\begin{quote}
\textit{Note: The authors attempted standardization of street names  
in an attempt to match what were obvious identical streets, deemed unequal 
due only to spelling and naming conventions. Those efforts were intended to match, for example, 
\textit{KANE PL} and \textit{KANE PLACE} as well as \textit{AVE P} and \textit{AVENUE P}. However, these 
efforts were  abandoned after they resulted in an increase in street name matches of less than 1\%.)}
\end{quote}

@@

    \item \texttt{cross\_street\_1/2} and \texttt{intersection\_street\_1/2}: 
    These two street pairs exhibit an 88\% match. Although we cannot 
    confirm the accuracy of either of these pairs, we recommend deleting the 
    \texttt{intersection\_street\_1/2} fields, acknowledging 
    the concomitant data loss.
\end{itemize}

\vspace{\baselineskip}
\noindent
Additionally, certain fields may not be useful for broad analysis
due to their sparse data population. This caveat applies to the fields 
below which are 99\% (or greater) blank.

\vspace{\baselineskip}
\noindent
\begin{quote}
\textit{We should note that even a field that is 99\% 
blank will still contain 64,000 non-blank rows in this 
large (6.4 million row) dataset.; certainly a non-trivia amount of data.}
\end{quote}

@@

The 311 SR Data Dictionary needs updating. We found several 
discrepancies between the Data Dictionary and the actual
data. Such errors could potentially mislead analytic efforts. Examples::

@@

\section*{Supplementary Material}
The NYC 311 R code used to produce these results, along 
with additional analytical findings, are available 
at \url{https://github.com/jun-yan/nyc311clean}. Reference 
data files (311 SR data, USPS zip codes) are available 
at \url{https://doi.org/10.6084/m9.figshare.c.7617311}.
